\Title

### 2011-03-08 masters_thesis/proposal/proposal.tex
Cuts to the proposal from Dr. Seppi and additions from Dr. Ventura

% Committee members
\committeechair{Kevin Seppi}
\committeemembera{Eric Ringger}
\committeememberb{David Embley}
%\committeememberc{}
%\committeememberd{}

@@

Particle swarm optimization (PSO) has previously been parallelized primarily by
distributing the computation corresponding to particles across multiple
processors.  In these approaches, additional processors can only add more
particles to the swarm.  However, in many cases this is not efficient when
scaled to very large swarm sizes (on very large clusters).  Current methods
cannot answer well the question, ``What do you do with 1000 processors when 50
or 100 particles is the most efficient swarm size?''  In this paper we attempt
to answer that question with a speculative approach to the parallelization of
PSO motivated by speculative execution in modern computer hardware.

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\citep{chu-2006-intelligent-parallel-pso,jin-2005-pso-antenna-designs,%
parsopoulos-2004-parallel-vector-evaluated-pso,%
schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
\citep{mostaghim-2006-multi-objective-pso-on-grids,%
venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{iters_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of iterations of the algorithm performed.}
  \label{fig:iters-sphere}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of function evaluations performed.  Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

In this project we consider PSO parallelization strategies for clusters of
hundreds or thousands of processors and functions for which a single evaluation
will take at least tens of seconds, but probably minutes or perhaps hours.  Our
purpose is to explore the question of what to do with a thousand processors
when 50 or 100 particles is the most efficient swarm size, and simply adding
particles results in only incremental improvement.  In our work we assume that
the researcher has some fixed number of processors that should be fully
utilized throughout the optimization process.

In order to solve this problem, we apply the concept of speculative
decomposition~\citep{grama-2003-intro-to-parallel-computing} to particle swarm
optimization, using extra processors to perform two iterations of PSO at the
same time.  Speculative decomposition is analogous to speculative execution
(also known as branch prediction), a technique commonly used in processors.
Modern processors, when faced with a branch on which they must wait (e.g., a
memory cache miss), guess which way the branch will go and start executing,
ensuring that any changes can be undone.  If the processor guesses right,
execution is much farther ahead than if it had idly waited on the memory
reference.  If it guesses wrong, execution restarts where it would have been
anyway.  Thus the processor speculates about future paths of execution in an
attempt to decrease overall processing time.

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach adapted
from speculative execution. We prove that the standard PSO equation can be
factored such that a set of speculative positions can be found which will
always include the position computed in the next iteration.  By computing the
value of the objective function for each of the speculative positions at the
same time the algorithm evaluates the objective function for the current
position, it is possible to know the objective function values for both the
current and the next iteration at the same time.  We demonstrate this principle
by implementation and show that it produces exactly the same results as
standard PSO, but two iterations at a time.  The resulting implementation runs
efficiently on large clusters where the number of processors is much larger
than a typical or reasonable number of particles, producing better results in
less ``wall-clock'' time.

@@

Changing the way neighbors are defined, usually called the ``topology,'' has a
significant effect on the performance of the algorithm.  In the Ring topology,
each particle has one neighbor to either side of it; in the Complete
topology\footnote{The Complete topology has often been unfortunately named Star
in the literature, which in graph theory refers to a completely different
topology.  Other names have also been used, including ``global topology'' and
gbest.  We use the graph theory term ``Complete'' in this paper.}, every
particle is a neighbor to every other
particle~\citep{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
particle is also a neighbor to itself in that its own position and value are
considered when updating the particle's neighborhood best, $\nbest$.  Thus with
$p$ particles, using the Ring topology each particle with index $i$ has three
neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
particle has $p$ neighbors.

In this paper we use these topologies as well as a parallel adaptation of the
Complete topology, called Random, that has been shown to approximate the
behavior of Complete with far less
communication~\citep{mcnabb-2009-large-particle-swarms}.  In the Random
topology, each particle randomly picks two other particles to share information
with at each iteration, along with itself.  Thus in both the Ring and the
Random topologies, all particles have three neighbors.

The ideal topology and swarm size for PSO depend on the objective function.
Researchers have devised various benchmark functions and have found that the
ideal topology for one function may perform very poorly for another function.
The No Free Lunch Theorems for Optimization show that this is true in
general---if an algorithm performs well on average for one class of functions
then it must do poorly on average for other
problems~\citep{wolpert-1997-nfl-for-optimization}.

An attempt to standardize PSO found that although a Complete swarm of 50
particles converged to the global optimum more quickly for many benchmark
functions than the same swarm size with a Ring topology, it was also more
likely to permaturely converge to local optima for other functions.  The study
found no significant improvement for any other swarm size between 20 and 100
and concluded with a recommendation to use a swarm of 50 particles with a Ring
topology as a starting point.  The authors acknowledged that choosing the ideal
topology requires thorough experimentation for the particular
problem~\citep{bratton-2007-defining-a-standard-for-pso}.  Other authors have
explained that a large swarm with a sparse topology propogates information
slowly~\citep{montes-de-oca-2009-frankensteins-pso}.

@@

\section{Related Work}
% 1 to 2 pages

There have been several parallel implementations of PSO presented in the
literature.  These implementations vary along two axes---what computation is
performed and how it is done.  How the computation is performed involves both
architectural decisions, such as a master-slave versus a fully distributed
architecture, and decisions about synchronicity.  What computation is performed
involves decisions about how to scale the PSO algorithm for parallel
implementations, such as the ratio of particles to processors or the topology
to use.

\subsection{Synchronous vs. Asynchronous PSO}

Synchronous parallel implementations of PSO reproduce the standard serial
algorithm exactly.  This approach was first described analytically by
\citet{belal-2004-parallel-models-for-pso} and first implemented by
\citet{schutte-2004-parallel-global-optimization-with-pso}.  In a typical
master-slave algorithm, the master assigns tasks to slave processors, and in
parallel PSO, each task consists primarily of a function evaluation.  Updating
the particle's position and value may also be included in the
task~\citep{belal-2004-parallel-models-for-pso}, or this work may be performed
in serial by the
master~\citep{schutte-2004-parallel-global-optimization-with-pso}.  Before
proceeding to the next iteration, particles communicate, and each particle
updates its neighborhood best.  Whether this communication step happens
sequentially on the master or in parallel, each particle must receive
communication from its neighbors before proceeding.  The benefits of the
synchronous PSO include its simplicity, repeatability, and comparability with
standard PSO, which may be essential in research applications.

Asynchronous parallel
PSO~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso} is a modification to the standard algorithm
which removes the synchronization point at the end of each iteration.  Instead,
particles iterate independently and communicate asynchronously.  In a typical
master-slave implementation of asynchronous parallel PSO, the master updates
each particle's personal best, neighborhood best, velocity, and position
immediately after receiving the function value from the slave processor.  Since
this update occurs while other particles are still being evaluated, it may use
information from the previous iteration for some
neighbors.\footnote{Asynchronous parallel PSO has been compared to the
``asynchronous updates'' variant of serial
PSO~\citep{koh-2006-parallel-asynchronous-pso}.  However, serial PSO with
asynchronous updates differs from standard PSO in that particles use newer
information, but asynchronous parallel PSO differs from standard PSO in that
particles use older information.} In a partially asynchronous implementation,
particles might wait for some but not all neighbors to complete before
proceeding~%
\citep{scriven-2008-asynchronous-pso-in-unreliable-distributed-environments}.
In some master-slave implementations, particles never get more than one
iteration ahead of
others~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.  However, in a fully distributed
implementation, particles might never wait for information, and one particle
could complete many more iterations than another
particle~\citep{scriven-2008-distributed-pso-using-peer-to-peer}.  The main
effect of asynchronous evaluation is that processors spend less time
idle---this trait is particularly valuable when processors are heterogeneous or
function evaluation times are
varied~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.  Asynchronous parallel PSO behaves
differently than the standard algorithm and may even produce different results
between runs.  Most reports conclude that asynchronous communication produces
similar numerical results to the standard algorithm, but the question has not
yet been thoroughly
addressed~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.

@@

\subsection{Refactoring the PSO Equations}

To perform two iterations at a time we must first refactor PSO such that the
determination of the value of the objective function is separate from the rest
of the computation.  For simplicity, this discussion will describe the case
where PSO is performing function minimization using the Ring topology.  In this
example, each particle has two neighbors, the ``right neighbor'' and ``left
neighbor,'' whose positions are represented as $\rightn$ and $\leftn$
respectively.  Though we will only describe the case of the Ring topology, this
method can easily be extended to arbitrary topologies.

@@

@@

\item March 7, 2011: Finish running experiments and incorporate the results
  into the thesis.

@@

\item March 29, 2011: submit thesis to second committee member for review.

>>>>

% Committee members
\committeechair{Kevin Seppi}
\committeemembera{Dan Ventura}
\committeememberb{David Embley}
%\committeememberc{}
%\committeememberd{}

@@

Particle swarm optimization (PSO) is an evolutionary algorithm that has proven
effective at solving complicated optimization problems.  Since its inception,
however, PSO has been run primarily in serial on a single machine.  With the
world's computing resources increasingly coming from large clusters of
processors, the parallelization of the algorithm is important to its continued
relevance.  PSO has previously been parallelized primarily by distributing the
computation corresponding to particles across multiple processors.  In these
approaches, additional processors can only add more particles to the swarm.
However, in many cases this is not efficient when scaled to very large swarm
sizes (on very large clusters).  Current methods cannot answer well the
question, ``What do you do with 1000 processors when 50 or 100 particles is the
most efficient swarm size?''  In this project we attempt to answer that
question with a speculative approach to the parallelization of PSO motivated by
speculative execution in modern computer hardware.

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\citep{chu-2006-intelligent-parallel-pso,jin-2005-pso-antenna-designs,%
parsopoulos-2004-parallel-vector-evaluated-pso,%
schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
\citep{mostaghim-2006-multi-objective-pso-on-grids,%
venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this project.

@@

\begin{figure}
  \centering
  \includegraphics[width=.4\columnwidth]{iters_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of iterations of the algorithm performed.}
  \label{fig:iters-sphere}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.4\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of function evaluations performed.  Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

In this project we will consider PSO parallelization strategies for clusters of
hundreds or thousands of processors and functions for which a single evaluation
will take at least tens of seconds, but probably minutes or perhaps hours.  Our
purpose is to explore the question of what to do with a thousand processors
when 50 or 100 particles is the most efficient swarm size, and simply adding
particles results in only incremental improvement.  In our work we assume that
the researcher has some fixed number of processors that should be fully
utilized throughout the optimization process.

In this project we will show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach adapted
from speculative execution. We prove that the standard PSO equation can be
factored such that a set of speculative positions can be found which will
always include the position computed in the next iteration.  By computing the
value of the objective function for each of the speculative positions at the
same time the algorithm evaluates the objective function for the current
position, it is possible to know the objective function values for both the
current and the next iteration at the same time.  We will demonstrate this
principle by implementation and show that it produces exactly the same results
as standard PSO, but two iterations at a time.  The resulting implementation
should run efficiently on large clusters where the number of processors is much
larger than a typical or reasonable number of particles, producing better
results in less ``wall-clock'' time.

@@

@@

\section{Related Work}
\label{sec:related}

There have been several parallel implementations of PSO presented in the
literature.  The improvements described in the literature come in two major
areas: innovations in implementation details and innovations in the use of
topology and swarm size to scale PSO to many processors.

There are many ways to parallelize the basic PSO algorithm.  The most
fundamental decision to make in parallel PSO is which parallel architecture to
use.  Several architectures have been proposed, including Master-Slave, fully
distributed (sometimes called ``diffusion''), and reformulating PSO into
Google's MapReduce framework~\citep{belal-2004-parallel-models-for-pso,
mcnabb-2007-parallel-pso-using-mapreduce}.  A somewhat orthogonal
implementation decision when parallelizing PSO is whether to have synchronous
communication or asynchronous communication.

@@

Our speculative methods require refactoring the PSO equations such that all
possible positions for each particle at iteration $t+1$ can be evaluated in
parallel along with the position of each particle at iteration $t$.  With some
careful bookkeeping, we can then piece together the results of iteration $t+1$
for each particle, thus using extra processors to evaluate two iterations of
the algorithm in the time it takes to evaluate the function once.  A wise
choice of topology limits the necessary speculative evaluations to seven per
particle.

To see the value of this refactoring, suppose that $1000$ processors are
available, and that the evaluation of the objective function takes one hour.
If we only want a swarm of $100$ particles, $900$ of the processors would be
sitting idle for an hour at every iteration, and it would take two hours to run
two iterations.  If instead we perform speculative evaluation, sending each of
the $7$ possible speculative positions of a particle to be computed at the same
time as its current position, we would use $800$ of the $1000$ processors and
perform two iterations in one hour.

In order to do two iterations at once, we must use 8 times as many processors
as there are particles in the swarm.  If these processors were not performing
speculative evaluation, they might instead be used for function evaluations
needed to support a larger swarm.  This raises the question of whether a swarm
of 100 particles doing twice as many iterations outperforms a swarm of 800
particles.  This project will seek to answer that question.

My honors thesis proposed and developed this basic approach to parallelization
in PSO.  This masters thesis will complete the picture of speculative methods
in particle swarm optimization.  First, we present a mathematical proof of
correctness of our methods, given in Section~\ref{sec:proof} as demonstration
that what we propose is in fact mathematically equivalent to standard
parallelization methods in PSO.  Second, we propose pruning the set of possible
speculative evaluations to better use available processors.  We motivate the
pruning by analyzing the statistics of which branches are commonly taken in
PSO, using those statistics to prune all but the most likely branches.
Further, once the idea of pruning is established, we view the set of all
possible speculative evaulations as an infinite tree, not confining ourselves
to speculating only about the next iteration.  We then experiment with
speculating down likely paths three or four iterations ahead, and show that
this by far outperforms simple one-iteration speculative evaluation.  And
lastly, as my honors thesis used only a few experiments on three benchmark
functions, we give a better picture of the performance of speculative
approaches with much more extensive experimentation.

\subsection{Refactoring the PSO Equations}
\label{sec:proof}

To perform two iterations at a time we must first refactor PSO such that the
determination of the value of the objective function is separate from the rest
of the computation.  For simplicity, this discussion will describe the case
where PSO is performing function minimization using the Ring topology.  In this
example, each particle has two neighbors, the ``right neighbor'' and ``left
neighbor,'' whose positions are represented as $\rightn$ and $\leftn$
respectively.  Though we will only describe the case of the Ring topology, this
method can easily be extended to arbitrary topologies (and in fact our
implementation of this algorithm works with arbitrary topologies; we simply
only discuss the Ring case here as that greatly simplifies our notation).

@@

For each function, we will use the topology that is generally considered to be
best for optimizing that particular function.  When the Complete topology is
considered best, we will instead use a Random topology that we developed that
mimicks the behavior of Complete while being more suited for parallel
computation~\cite{mcnabb-2009-large-particle-swarms}.

The functions we will use in our comparisons are these common benchmark
functions: Bohachevsky, Griewank, Rastrigin, Sphere, and Schwefel 2.21.  We
will run each function with 20, 50, and 500 dimensions to test the scalability
of our methods up to higher dimensions.  We will also test our methods on a
more typical ``real-world'' problem, that of training a computationally
intensive machine learning algorithm with a large amount of data.  We will
compare our speculative methods to previous parallelization techniques,
including a naive parallelization of just increasing the swarm size, and a
subswarms implementation that uses several smaller, independent swarms.

@@

\item March 11, 2011: Finish running experiments and incorporate the results
  into the thesis.

@@

\item March 22, 2011: submit thesis to second committee member for review.
